---
title: "Lab 3"
author: "Simon Jorstedt"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(ade4)
library(ape)
library(dplyr)
library(magrittr)
library(mvMORPH)
library(mvSLOUCH)

data(carni70)
```

# Q 2
## Q 2.1
*Explore the data set and report what can be found in it. Provide some plots*

The `carni70` dataset describes the phylogeny, as well as geographic range (km) and body size (kg) of 70 carnivorous animals. In Figure 1 and 2 below we find that a majority of the animals are relatively small/light (under 10kg), with some outliers in the 20-50 kg range, and just two in the range of 200-260 kg range. Most of the animals belong to either a cluster with low range (up to about 7 km), or a cluster with longer range (around 12-27 km). In Figure 3 we see a phylogenetic tree of the animals

```{r}
# Plot the range vs the size
plot(carni70$tab$size, carni70$tab$range,
     xlab="Body size (kg)", 
     ylab="Geographical range (km)",
     main="Fig. 1: Geographical range vs body size")


# Plot range vs size of small animals
small_carni <- carni70$tab %>% 
  filter(size < 50)

plot(small_carni$size, small_carni$range,
     xlab="Body size (kg)", 
     ylab="Geographical range (km)",
     main="Fig. 2: Geographical range vs body size of small animals (< 50kg)",
     xlim = c(0,50))


# Plot the phylogeny
carni_tree <- ape::read.tree(text = carni70$tre)
plot(carni_tree, type="fan",
     main="Fig. 3: Species phylogeny")

# Change carni70 names, since names are separated with "." in the tree
rownames(carni70$tab) <- gsub("_", replacement = ".", rownames(carni70$tab))
```

## Q 2.2
*Analyze the two traits (size and range) with a number of different phylogenetic comparative models. Look at the R packages ape, mvMORPH, mvSLOUCH, ouch and slouch (the last one is NOT on CRAN and can be downloaded from https://github.com/kopperud/slouch). We will analyse the data under five different models, described in the code chunk below.*

Below we train the models, and then take some interesting results (mainly the loglikelihood and AIC scores) and display them in a dataframe. We find that both the Brownian motion models are practically  the same, while the two Ornstein-Uhlenbeck models are slightly better, and about as good as the mixed model. There is a bit of random variation in the convergences, so it is unclear and ambiguous which model is the best, but it will be either the dependent OU model, or the mixed model.

```{r}
# Both traits evolving as independent Brownian motions
bm_indep <- mvMORPH::mvBM(carni_tree,
                          data = carni70$tab,
                          model="BM1",
                          param = list(constaint="diagonal"))

# The traits evolve as a correlated Brownian motion
bm_dep <- mvMORPH::mvBM(carni_tree,
                        data = carni70$tab, 
                        model="BM1")

# Both traits evolving as independent Ornstein-Uhlenbeck processes
ou_indep <- mvMORPH::mvOU(carni_tree,
                          data=carni70$tab,
                          model="OU1",
                          param = list(constaint="diagonal"))

# Traits evolving as bivariate Ornstein-Uhlenbeck process
ou_dep <- mvMORPH::mvOU(carni_tree,
                        data=carni70$tab,
                        model="OU1")

mixed_mod <- mvSLOUCH::mvslouchModel(carni_tree,
                        as.matrix(carni70$tab[c(2,1)]),
                        kY=1)

# mvSLOUCH is an... interesting package
if (mixed_mod$MaxLikFound == "Same as final found"){
  mixed_mod$MaxLikFound <- mixed_mod$FinalFound 
}

# Format results of the analyses
list("BM indep"=c(bm_indep$LogLik,
                  bm_indep$AIC,
                  bm_indep$AICc,
                  bm_indep$convergence,
                  bm_indep$hess.values),
     
     "BM dep"=c(bm_dep$LogLik,
                bm_dep$AIC,
                bm_dep$AICc,
                bm_dep$convergence,
                bm_dep$hess.values),
     
     "OU indep"=c(ou_indep$LogLik,
                  ou_indep$AIC,
                  ou_indep$AICc,
                  ou_indep$convergence,
                  ou_indep$hess.values),
     
     "OU dep"=c(ou_dep$LogLik,
                ou_dep$AIC,
                ou_dep$AICc,
                ou_dep$convergence,
                ou_dep$hess.values),
     
     "Mixed"=c(mixed_mod$MaxLikFound$LogLik,
               mixed_mod$MaxLikFound$ParamSummary$aic,
               mixed_mod$MaxLikFound$ParamSummary$aic.c,
               NA,
               NA)) %>%
  do.call(rbind, .) %>%
  set_colnames(c("LogLikelihood", "AIC", "AICc", "convergence", "hess.values"))
```




